# Some statistical tests in R {#tests}

*Pimonrat Tiansawat (Chiang Mai University)*

Updated: `r format(Sys.Date(), '%B, %Y')`

**Duration:** 8 sessions

## Learning Outcomes

Students should be able to:

1.  demonstrate how to run some statistical tests and visualize data.

## Data Files for This Lab

- Data in the folder[Click here to download the file](https://o365cmu-my.sharepoint.com/:f:/g/personal/pimonrat_t_cmu_ac_th/IgCK_zguJAD2RLvlfp9br6QHAQK6NhRqHerDwNXCqPj4biQ?e=CR0iBz)

After downloading keep the data folder in your working directory.

```{r setup, include = F}
knitr::opts_chunk$set(echo = TRUE)
```

## Parametric tests

### z test

We would like to test whether River A is cleaner than River B. We have DO data from both rivers.

```{r Z-test, echo = TRUE, eval = FALSE}
#########################
### z-test
#########################
if(!require(BSDA)){install.packages("BSDA")}
library(BSDA)

dat <- read.csv('data/z-test.csv', header = T)

boxplot(dat[1:30, 2], dat[31:60, 2], xlab = "River", ylab = "DO (mg/l)", names = c("A", "B"))

z.test(x = dat[1:30, 2], 
       y = dat[31:60, 2], 
       alternative = 'greater',
       sigma.x = sd(dat[1:30, 2]),
       sigma.y = sd(dat[31:60, 2]))

```

What is our conclusion?

### t test

We would like to test whether the Toxic A (treatment 1) has an effect on the kidney weight of mice in comparison to the control (treatment 2).

```{r t-test, echo = TRUE, eval = FALSE}
#########################
### t-test
#########################

dat <- read.csv('data/t-test.csv', header = T)
dat$Treatment <- as.factor(dat$Treatment)

boxplot(Weight ~ Treatment, data = dat)
t.test(data = dat,
       Weight ~ Treatment,
       alternative = c("greater"),
       mu = 0, var.equal = FALSE,
       conf.level = 0.95)
```

What is the conclusion?

### F test

We would like to test whether the variation of gibbon weight are equal between two forests.

```{r F-test, echo = TRUE, eval=FALSE}
#########################
### f test 
#########################

dat <- read.csv('data/F-test.csv', header = T)

boxplot(gibbon_weight ~ forest, data = dat)
dat$forest<- as.factor(dat$forest)
var.test(gibbon_weight ~ forest, data = dat, 
         alternative = "two.sided")
```

Then we can choose an appropriate t test to test the mean weight differences. 

### Paired t test

We would like to test whether two teaching methods (n = 10) give the same score outcome at significant level = 0.01.

```{r Paired t-test, echo = TRUE, eval = FALSE}
#########################
### Paired t-test
#########################
dat <- read.csv('data/pair-test.csv', header = T)
dat$teach <- as.factor(dat$teach)

boxplot(score ~ teach, data = dat)
t.test(x = dat[1:10, 2], 
       y = dat[11:20, 2],
       alternative = c("greater"), paired = T,
       mu = 0, var.equal = FALSE,
       conf.level = 0.99)

### Check an assumption of paired t tests
d <- with(dat, 
          score[teach == "1"] - score[teach == "2"])
hist(d)
shapiro.test(d)

```

Let's visualize paired data

```{r Paired plot, echo = TRUE, eval=FALSE}
if(!require(PairedData)){install.packages("PairedData")}
library(PairedData)

teach1 <- subset(dat,  teach == "1", score,
                 drop = TRUE)

teach2 <- subset(dat,  teach == "2", score,
                drop = TRUE)
pd <- paired(teach1, teach2)
plot(pd, type = "profile")

```

### One-way ANOVA

We determine whether the four methods yield the different amount of phosphorus in grasses.

```{r oneway, echo = TRUE, eval = FALSE}

dat <- read.csv('data/CRD.csv', header = T)
dat$Method <- as.factor(dat$Method)

plot(Phos ~ Method, data = dat)

library(dplyr)
summary.crd <- dat %>%
    group_by(Method) %>%
    summarise(mean = mean(Phos),
              sd = sd(Phos),
              max = max(Phos),
              min = min(Phos),
              SE = sd(Phos)/sqrt(length(Phos) - 1))

### Bar plot with error bars
if(!require(ggplot2)){install.packages("ggplot2")}
library(ggplot2)

ggplot(data = summary.crd,
       aes(x = Method, y = mean, fill = Method)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - SE, ymax = mean + SE),
              width = 0.2, # Width of the error bars
              position = position_dodge(0.9))

#### First, test for equal variance
fligner.test(Phos ~ Method, data = dat)                   

#### ANOVA
model <- aov(Phos ~ Method, data = dat) 
summary(model)

#### Check assumptions of normality of residuals
shapiro.test(model$residuals)
hist(model$residuals)

par(mfrow = c(2, 2)) 
plot(model) 

#### See effect
model.tables(model, se = T) # compare to the overall mean
summary.lm(model) #compare to Method 1

#### Multiple comparisons
if(!require(mulcomp)){install.packages("multcomp")}
library(multcomp)

## See pairwise comparisons
summary(glht(model, linfct = mcp(Method = 'Tukey')))

## Display letters
cld(glht(model, linfct = mcp(Method = 'Tukey')))
```

If it turn out that the assumption of equal variance is not met, we can use *oneway.test* function to do Welch's ANOVA that does not assume equal variance.

```{r oneway unequal var, echo = TRUE, eval = FALSE}
oneway.test(Phos ~ Method, data = dat, var.equal = FALSE)

```

### One-way ANOVA with blocking factors

Plants got four different hormones, and planted in sites with three levels of different moisture availability.
We would like to test whether the method yield the different of chemical of interest.

```{r oneway with block, echo = TRUE, eval=FALSE}

dat <- read.csv('data/RCB.csv', header = T)
dat$Hormone <- as.factor(dat$Hormone)
dat$Block <- as.factor(dat$Block)

par(mfrow = c(1,2))
plot(Substance ~ Hormone + Block, data = dat)

library(dplyr)
summary.rcb <- dat %>%
    group_by(Hormone) %>%
    summarise(mean = mean(Substance),
              sd = sd(Substance),
              max = max(Substance),
              min = min(Substance),
              SE = sd(Substance)/sqrt(length(Substance) - 1))
### Interaction plot
interaction.plot(x.factor     = dat$Hormone,
                 trace.factor = dat$Block,
                 response     = dat$Substance,
                 fun = mean,
                 type = "b",
                 col = c("black","red","green"),  ### Colors for levels of trace var.
                 pch = c(19, 17, 15),             ### Symbols for levels of trace var.
                 fixed = TRUE,                    ### Order by factor order in data
                 leg.bty = "o")

### Bar plot with error bars                               
ggplot(data = summary.rcb,
       aes(x = Hormone, y = mean, fill = Hormone)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = mean - SE, ymax = mean + SE),
              width = 0.2, # Width of the error bars
              position = position_dodge(0.9))

ggplot(data = dat,
       aes(x = Hormone, y = Substance, color = Block)) +
  geom_boxplot()

#### First, test for equal variance
fligner.test(Substance ~ Hormone, data = dat)                   
#### ANOVA
model <- aov(Substance ~ Hormone + Block, data = dat) 
summary(model)

#### Check assumptions of normality of residuals
shapiro.test(model$residuals)
hist(model$residuals)

par(mfrow = c(2, 2)) 
plot(model) 

#### See effect
model.tables(model, se = T) # compare to the overall mean
summary.lm(model) #compare to Method 1

#### Multiple comparisons
if(!require(mulcomp)){install.packages("multcomp")}
library(multcomp)

## See pairwise comparisons
summary(glht(model, linfct = mcp(Hormone = 'Tukey')))
summary(glht(model, linfct = mcp(Block = 'Tukey')))

## Display letters
cld(glht(model, linfct = mcp(Block = 'Tukey')))

```

### Two-way ANOVA

There are two plant cultivars, received three different fertilizers.
Blocking is the three plots, located at different distance from the water sources.
The dependent variable is yield (kg/Rai)

```{r twoway, echo = TRUE, eval=FALSE}

dat <- read.csv('data/FactorialRCB.csv', header = T)
dat$Cultivar <- as.factor(dat$Cultivar)
dat$Fertilizer <- as.factor(dat$Fertilizer)
dat$Rep <- as.factor(dat$Rep)
dat$Combination<- as.factor(dat$Combination)

par(mfrow = c(1,2))
plot(Biomass ~ Cultivar + Fertilizer, data = dat)

library(dplyr)
summary.fac <- dat %>%
    group_by(Fertilizer) %>%
      summarise(mean = mean(Biomass),
              sd = sd(Biomass),
              max = max(Biomass),
              min = min(Biomass),
              SE = sd(Biomass)/sqrt(length(Biomass) - 1))

### boxplot
ggplot(data = dat,
       aes(x = Cultivar, y = Biomass,
           fill = Fertilizer)) +
  geom_boxplot()

### Interaction plot
interaction.plot(x.factor     = dat$Fertilizer,
                 trace.factor = dat$Cultivar,
                 response     = dat$Biomass,
                 fun = mean,
                 type = "b",
                 col = c("black","red","green"),  ### Colors for levels of trace var.
                 pch = c(19, 17, 15),             ### Symbols for levels of trace var.
                 fixed = TRUE,                    ### Order by factor order in data
                 leg.bty = "o")

#### First, test for equal variance of all combinations
fligner.test(Biomass ~ Combination, data = dat)                   
#### ANOVA model with interaction terms
model <- aov(Biomass ~ Cultivar * Fertilizer, data = dat) 
summary(model)

final.model <- step(model)
summary(final.model)

#### Check assumptions of normality of residuals
shapiro.test(final.model$residuals)
hist(final.model$residuals)

par(mfrow = c(2, 2)) 
plot(model) 

#### See effect
model.tables(final.model, se = T) # compare to the overall mean
summary.lm(final.model) #compare to Method 1

#### Multiple comparisons
if(!require(emmeans)){install.packages("emmeans")}
library(emmeans)

marginal = emmeans(final.model,
                   ~ Fertilizer:Cultivar)

pairs(marginal,
      adjust="tukey")

library(multcomp)
## compare Cultivar
cld(glht(model, linfct = mcp(Cultivar = 'Tukey')))

## compare Fertilizer
cld(glht(model, linfct = mcp(Fertilizer = 'Tukey')))

## compare all combinations
cld(marginal,
    alpha = 0.05,
    Letters = letters,    ### Use lower-case letters for .group
    adjust="tukey") 

```

### Chi-square test

Test whether the ratio of red to white flowers is not 3 to 1 We observe 21 red and 19 white flowers.
The expected ratio is 3:1.

```{r chi-square prob, echo = TRUE, eval=FALSE}

observed <- c(21, 19) ## provide a vector of count
expected <- c(3/4, 1/4) ## provide a vector of expected prob.
chisq.test(observed, p = expected)

```

### Another type of chi-square test

H0 having cancer is independent from smoking.
Ha having cancer is dependent from smoking.

```{r chi-square indep, echo = TRUE, eval=FALSE}

dat <- read.csv('data/Chi-square.csv', header = T)

## organize data frame
rownames(dat) <- dat$X
dat <- dat[,-1]

chisq.test(dat) ## run the test

```

## Correlation analysis

We will test the correlation between the wing length and tail length.

```{r cor.test, echo = TRUE, eval=FALSE}

dat <- read.csv('data/Correlation.csv', header = T)

## organize data frame
dat <- dat[,-1]

cor.test(dat$wing.length.cm, dat$tail.length.cm)
```

Correlation coefficient is 0.85.
and R-squared is 0.72.
There is correlation between the wing and tail length.
Let's see a plot.

```{r plot, echo = T, eval = F, fig.show = "hide"}
library(ggplot2)
plot <- ggplot(dat, 
               aes(x = wing.length.cm,
                   y = tail.length.cm))

plot +
  geom_point() +
  xlab("\n Wing length (cm)") +
  ylab("Tail length (cm)\n")
```

## Regression analysis

```{r package installing, echo = T, eval = FALSE}
if(!require(gvlma)){install.packages("gvlma")}
```

Let's get the data set.
We want to find the relationship between the age (days) and the length of wings (cm) of a species of birds

```{r import, echo = T, eval = T, results ='hide'}
### Import data
reg.dat <- read.csv("data/Regression.csv", header = T)

### assign right types of data to the variables
reg.dat$age_day <- as.numeric(reg.dat$age_day)

### see the data
reg.dat
```

Now let's visualize the two variable using a scatter plot

```{r reg.plot, echo = T, eval = F, fig.show = "hide"}
library(ggplot2)
plot.reg <- ggplot(reg.dat, 
               aes(x = age_day,
                   y = wing_length_cm))

plot.reg +
  geom_point() +
  xlab("\nAge (days)") +
  ylab("Wing length (cm)\n")
```

Now, let's run linear regression model.
We will evaluate how well the model fit and get the linear equation.

Y = β₀ + β₁X + ε

-   Y is wing length
-   X is age
-   β₀ is intercept and
-   β1 is slope
-   ε is error term We use the *set.seed()* function to guarantee that the same random values are produced each time you run the code.

```{r reg.model1, echo = T, results = 'hide'}
##################################################
### Run a regression model
### and run a function to get
### overall p value
##################################################
set.seed(1234)
model <- lm(wing_length_cm ~ age_day,
            data = reg.dat)

summary(model)

overall_p <- function(my_model) {
    f <- summary(my_model)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}

overall_p(model)

anova(model)

intercept <- coef(model)[1]
slope <- coef(model)[2]
```

The summary of the model shows the significant parameter estimates of intercept = `r intercept` and slope = `r slope`.
This means that the linear equation is

`r model$terms[[2]]` = `r intercept` + `r slope` \* (`r model$terms[[3]]`) + ε

Now, we will evaluate the model to see if the model met the assumptions.
First, let's check if the residuals are normally distributed with mean of 0.

```{r model.eval1, echo = F, eval = F, results = 'hide', fig.keep = 1}

hist(model$residuals)
abline(v = mean(model$residuals))
qqnorm(model$residuals)
qqline(model$residuals, lwd = 2)
shapiro.test(model$residuals)
```

Second, let's see a set of four diagnostic plots.
We can check for heteroscedasticity using a fitted value vs residual plot.

```{r eval.plot, echo = F, eval = F}
par(mfrow=c(2,2)) 
plot(model)
```

See the upper left plot.
It does not look scattered.
Also, see the Q-Q residuals plot.
The residuals are not quite normally distributed.

In addition, there is a package that can help you check the assumptions of linear regression.
Let's try `gvlma()` function from `gvlma` package.

```{r gvlma, echo = F, eval = F, results = 'asis'}
library(gvlma)
gvlma::gvlma(model)

```

It seems like we passed the assumptions.
The linear regression is all right.

```{r add lm line, echo = T, eval = F, fig.show = "hide"}
plot.reg +
  geom_point() +
  xlab("\nAge (days)") +
  ylab("Wing length (cm)\n")+
  geom_abline(intercept = intercept, slope = slope,
              color = "purple")
```

## Nonparametric tests

### Mann-Whitney test or Wilcoxon Rank-Sum test

We would like to test whether the blood Hb of the animals exposed to CdO differ from that of the animals in a control group.
In R, Mann-Whitney U test uses the function wilcoxon.test().

```{r Mann, echo = TRUE, eval=FALSE}

#########################
### Import data sets
#########################
dat <- read.csv('data/Mann-Whitney.csv', header = T)
dat$Group <- as.factor(dat$Group)

boxplot(data ~ Group, data = dat)

### test
wilcox.test(data ~ Group, data = dat, 
            alternative = "greater")
```

### Wilcoxon Signed Rank Test

It is an alternative to the paired t-test.

```{r wilcoxon, echo = TRUE, eval=FALSE}

#########################
### Import data sets
#########################
dat <- read.csv('data/Wilcoxon.csv', header = T)
dat$Hindleg_length <- as.numeric(dat$Hindleg_length)
dat$Foreleg_length <- as.numeric(dat$Foreleg_length)

boxplot(dat$Foreleg_length, dat$Hindleg_length,
        ylab = "Leg length (cm)",
        names = c('Foreleg', 'Hind leg'))

### test
wilcox.test(dat$Foreleg_length, dat$Hindleg_length,
            alternative = "less", paired = TRUE)
```

### McNemar test for paired data

We would like to test whether the acid prodcution of microbes after fructose addition differs from that before adding fructose.

```{r McNemar, echo = TRUE, eval=FALSE}

#########################
### Import data sets
#########################
dat <- read.csv('data/Mc-Nemar.csv', header = T)

table(dat$before, dat$after)

### test
mcnemar.test(dat$before, dat$after)
```

McNemar test is for binary data.
For continuous data and ordinal data, Friedman test may be appropriate.

### Friedman's test

We would like to test whether the four stimulants have effects on rat behaviors.

```{r Friedman, echo = TRUE, eval=FALSE}

#########################
### Import data sets
#########################
dat <- read.csv('data/Friedman.csv', header = T)
dat$treatment <- as.factor(dat$treatment)
dat$batch <- as.factor(dat$batch)

boxplot(rat ~ treatment, data = dat, ylab = "Rat behaviour")

### test
friedman.test(rat ~ treatment | batch, data = dat)

### post hoc test
### p.adjust.methods include c("holm", "hochberg", "hommel", "bonferroni", "BH", "BY", "fdr", "none"). See the explanation in p.adjust function.
pairwise.wilcox.test(dat$rat, dat$treatment, p.adj = "bonf")
```

### Kruskal-Wallis test

It is an alternative to the one-way ANOVA.
Here we would like to test whether the drug effect the reaction time of three groups of animals.

```{r kruskal, echo = TRUE, eval=FALSE}

#########################
### Import data sets
#########################
dat <- read.csv('data/Kruskal.csv', header = T)
dat$Group <- as.factor(dat$Group)

boxplot(reaction ~ Group, data = dat, ylab = "Reaction time (min.)")

### test
kruskal.test(reaction ~ Group, data = dat)

### post hoc test
pairwise.wilcox.test(dat$reaction, dat$Group, p.adj = "bonf")
```

Now let's try again with the data that we did one way-ANOVA

```{r kruskal2, echo = TRUE, eval=FALSE}

#########################
### Import data sets
#########################
dat <- read.csv('data/CRD.csv', header = T)
dat$Method <- as.factor(dat$Method)

boxplot(Phos ~ Method, data = dat, ylab = "Amount of phosphate")

### test
kruskal.test(Phos ~ Method, data = dat)

### post hoc test
pairwise.wilcox.test(dat$Phos, dat$Method, p.adj = "holm")
```

### Cochran Q test

It is used to see if the proportion of successes is equal across three or more groups.
Here we would like to test whether the three treatments result in the different outcome.

```{r Cochrans, echo = TRUE, eval=FALSE}

### Install the required package
if(!require(RVAideMemoire)){install.packages("RVAideMemoire")}
library(RVAideMemoire)
#########################
### Import data sets
#########################
dat <- read.csv('data/CochranQ.csv', header = T)

dat$id <- as.factor(dat$id)
dat$treatment <- as.factor(dat$treatment)

table <- xtabs( ~ response + treatment,
               data = dat)

barplot(table,
        beside = TRUE,
        legend = TRUE,
        args.legend = list(x = "top", title = "response",
                           horiz = TRUE, ### location
                           bty = "n",
                           inset=c(0, -0.3)))

## test: It comes with pairwise comparisons, but it is not recommended.
cochran.qtest(response ~ treatment | id,
              data = dat)

## Post hoc with McNemar test
if(!require(rcompanion)){install.packages("rcompanion")}
library(rcompanion)

pair.compare <- pairwiseMcnemar(response ~ treatment | id,
                     data   = dat,
                     test   = "permutation",
                     method = "fdr",
                     digits = 3)

### see compact letter display
cldList(p.adjust ~ Comparison,
        data       = pair.compare$Pairwise,
        threshold  = 0.05)
```
