---
title: "Lab Session 14-15" 
subtitle: "Multivariate analyses in ecological and environmental studies"
author: "Pimonrat Tiansawat"
date: "Last updated on `r format(Sys.time(), '%d %B, %Y')`"
output:
  prettydoc::html_pretty:
    theme: cayman
    toc: true # Enables the table of contents
    toc_depth: 2  # Sets the depth of TOC (optional)
    smooth_scroll: true # Enables smooth scrolling for TOC links (optional)
  word_document:
    toc: false
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
    toc: false
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Learning Outcomes

Students should be able to:

1.  Give some examples of multivariate data analyses.
2.  Search for journal articles and information on related multivariate statistics and summarize key procedure of the related analyses. 

There are many multivariate data analyses. In this lab, three multivariate ordination techniques are introduced - Principal component analysis, cluster analysis and multidimensional scaling.

# Principal component analysis (PCA)

PCA is one of the most common multivariate data analyses. In R, PCA can be done with *princomp* and *prcomp* (both from stats package).

The *princomp* uses spectral decomposition which examines the covariances/correlations between variables, while the *prcomp* use singular value decomposition which examines the covariances/correlations between individuals. According to R help, the singular value decomposition method has better numerical accuracy, so prcomp is preferable (sthda.com). 

If you have not done so, install some packages.

```{r package installing, echo = T, eval = FALSE}
install.packages("GGally")
install.packages("ggplot2")
```

Then let's import a data set, specify the appropriate data types.

``` {r importing, echo = TRUE, eval = TRUE}
#########################
### Import data sets
#########################
seed <- read.csv('data/seed3spp.csv', header = T)

### assign right types of data to the variables
seed$genus <- as.factor(seed$genus)
seed$epithet <- as.factor(seed$epithet)

## set row names for later visualization 
rownames(seed) <- paste(seed$genus, seed$no) 

head(seed) ### seed few rows
summary(seed)

```

Here we have data of seed traits of three tree species. The number of seeds are 20 each. The data contain 60 rows and 11 columns. Let's look at the data set and see what trait data we have and see the correlation between each seed variable. 

```{r explore, echo = T, eval = T, fig.show = "hide"}
## use GGally package
library(ggplot2)
library(GGally)

## use data from column 7:11 - volume to moisture
ggpairs(seed,          
        columns = 7:11) # selected columns
```

The variables are correlated highly. Let's do PCA.

```{r pca, echo = TRUE, eval = FALSE}
### PCA

seed.pca <- prcomp(seed[,-c(1:3)], scale = T) ## use correlation matrix
seed.pca$x ## see scores of each sample

plot(seed.pca$x, pch = as.numeric(seed$genus))

#Interpret the results
summary(seed.pca)

screeplot(seed.pca)
screeplot(seed.pca, type = 'l') ## same meaning - it's a line now

seed.pca$rotation ## see the loadings

biplot(seed.pca) ## overlay the loading with score plot

```

Let's try different packages to make a better plot. See the contribution of each variable.

```{r better plot1, echo = TRUE, eval = FALSE}

install.packages("factoextra")
library(factoextra)

fviz_pca_biplot(seed.pca,
                label = "var",
                col.ind = "black",
                col.var = "contrib",
                gradient.cols = c("blue","green","red"))

fviz_contrib(seed.pca, 'var', axes = 1, xtickslab.rt = 45)

fviz_contrib(seed.pca, 'var', axes = 2, xtickslab.rt = 45)
```

Display species on the plot.

```{r better plot2, echo = TRUE, eval = FALSE}
fviz_pca_biplot(seed.pca,
                label = "var",
                habillage = seed$genus,
                addEllipses = TRUE,
                ellipse.level = 0.95,
                ggtheme = theme_minimal())
```

# Clustering with PCA scores

There are many algorithms to do cluster analyses. Here we will use K-means clustering. K-means is a popular centroid-based clustering algorithm used to group data into clusters based on similarity. Our objective is to group based on seed traits. 

```{r clustering, echo = TRUE, eval = FALSE}

# Extract PC1 and PC2 scores
pca.scores <- data.frame(seed.pca$x[, 1:2]) 

## determine number of clusters
fviz_nbclust(
  pca.scores,
  FUNcluster = kmeans,
  method = 'silhouette', k.max = 5
)
# Run k-means clustering with the determined number of clusters.

set.seed(123)
kmeans.clust <- kmeans(pca.scores, centers = 2)
kmeans.clust

### Visualization of clusters
fviz_pca_ind(seed.pca,
             habillage = kmeans.clust$cluster,
             repel = TRUE,
             addEllipses = TRUE,
             ellipse.type = "convex",
             labelsize = 2)

fviz_pca_biplot(seed.pca, 
                col.var = "black", 
                alpha.var = 0.6,
                label = "all",
                habillage = kmeans.clust$cluster,
                repel = TRUE,
                addEllipses = TRUE,
                ellipse.type = "convex", labelsize = 2)
```

# Non-metric multidimensional scaling (NDMS)

Multidimensional scaling is visualization technique based on the similarity of individual in a data set. There are five types of multidimensional scaling. In this lab, we will learn to visualize similarity of seeds based on their traits.


```{r nmds, echo = TRUE, eval = FALSE}

# packages
if (!require("vegan")) {install.packages("vegan")}
library(vegan)

seed.nmds <- metaMDS(seed[,-c(1:3)], # Our community-by-species matrix
                     k = 2) # The number of reduced dimensions

stressplot(seed.nmds)

plot(seed.nmds)
ordiplot(seed.nmds, type = "n")
orditorp(seed.nmds, display = "sites", cex = 0.3, air = 0.01)
orditorp(seed.nmds, display = "species", cex = 0.5, air = 0.01, col = "brown")

ordiplot(seed.nmds, type = "n")
orditorp(seed.nmds,display = "sites",col=c(rep("darkgreen",20),rep("blue",20), rep("red", 20)), air = 0.01, cex = 0.3)

ordihull(seed.nmds, groups = seed$genus, draw = "polygon", col = "grey90",
         label = F)

```
 

# Exercise

**Submission:** A PDF file containing your answers.
Your answers must contain **R codes** that you wrote and **the outputs** you get.
Your codes and outputs can be text and/or screen capture.

**Full score:** 10

## Questions/Directions

**1**.  Find a journal article (of your own interest) that has multivariate data analysis.

**2**.  Identify what research questions the author(s) want to answer.  

**3**.  Identify what the data used in the article are.

**4**.  Identify what models or analyses the author(s) used in the study 

**5**.  Search from external sources about the models/ analyses and summarize in your own words what the advantages and disadvantages of the models/ analyses are.

**6**.  Summarize in your own words what the procedures to run the analyses are.

**7**.  Include the list of references you used.

------------------------------------------------------------------------

## Grading Rubric

| Question | Score = 2 | Score = 1 | Score = 0 |
|------------------|------------------|------------------|------------------|
| 1 | - | The journal has at least one multivariate data analysis | The journal has NO multivariate data analysis |
| 2 | - | Correctly identify | Incorrectly identify or the answer is not included |
| 3 | - | Correctly identify | Incorrectly identify or the answer is not included |
| 4 | - | Correctly identify | Incorrectly identify or the answer is not included |
| 5 | Correct summary | Partial correct summary  | Incorrect summary or the answer is not included |
| 6 | Correct summary | Partial correct summary  | Incorrect summary or the answer is not included |
| 7 | - | Correct give the reference list | The reference list is not included |

------------------------------------------------------------------------

**End of Lab**
